New reduction techniques in commutative algebra driven by logical methods

### 2018-12-06 slides-padova2018.tex
Notes slides finished

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\CC}{\mathbb{C}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LocRing}{\mathrm{LocRing}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ull}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
\newcommand{\Ll}{\vcentcolon\!\Longleftrightarrow}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}{\vdash_{\!\!\!\vec x}}

@@

  \visible<4>{
    \begin{changemargin}{-2.0em}{-0.5em}
      \begin{itemize}
        \item \ \\[-1.2em]\mbox{For any reduced ring~$A$, there is a ring~$A^\sim$ in a certain topos with}
        \[ \models \bigl(\forall x\?A^\sim\_ \neg(\exists y\?A^\sim\_ xy = 1) \Rightarrow x = 0\bigr). \]

@@

  \jnote{1}{The two displayed statements are trivial for fields. It is
  therefore natural to try to reduce the general situation to the field
  situation.}

@@

  Using the new reduction technique, there is a short (one-paragraph) and
  conceptual proof of Grothendieck's generic freeness lemma. It is constructive
  as a bonus; and if desired, one can unwind the resulting proof to obtain a
  constructive proof which doesn't reference topos theory. The proof obtained
  in this way is still an improvement on the previously known proofs, requiring
  no advanced prerequisites in commutative algebra, and takes
  \fixedhref{https://arxiv.org/abs/1807.01231}{about a page}.}
\end{frame}

@@

  \jnote{1}{In topos theory, we have lots of experience of ``changing
  universes'' in order to ``force'' some statements to be true. However,
  because the field condition we are aiming at is not a ``geometric sequent'', these
  techniques do not work here. Hence we try to take it more slowly and devise a
  semantics which forces the given ring only to be local.

  The key insight is that \emph{locally} (in the sense of topology/geometry),
  any ring is a local ring. That is, we may pretend that any given ring is
  local if we are prepared to pass to numerous localizations during the course
  of an argument. The semantics displayed on the next slide manages this
  localization-juggling for us.}
\end{frame}

@@

  Irrespective of whether~$A$ is a local ring, its mirror
  image~$A^\sim$ is always a local ring (that is, the axioms of what it means
  to be a local ring hold under the translation rules specified by the
  semantics). A basic application of this forcing model are local-to-global
  principles. For instance:
  \begin{itemize}\justifying
    \item The statement ``the kernel of a surjective matrix over a local ring
    is finite free'' admits a constructive proof. It therefore holds
    for~$A^\sim$. Its external meaning is that the kernel of a surjective
    matrix~$M$ over~$A$ is finite locally free (there exists a partition~$1 =
    f_1 + \cdots + f_n$ such that for each~$i$, the localized
    module~$(\operatorname{ker} M)[f_i^{-1}]$ is finite free).

@@

@@

    \begin{column}{0.50\textwidth}
      \small\justifying
      For a fixed ring~$R$, the localization $A' \defeq A[S^{-1}]$ with $S \defeq
      \alpha^{-1}[R^\times]$ would do the job.
      \medskip

@@

  Prime filter:\, $0 \not\in \fff$;\,\,
  $x+y \in \fff \Rightarrow x \in \fff \vee y \in \fff$;
  \hspace*{0.5pt}\,\,
  $1 \in \fff$;\,
  $xy \in \fff \Leftrightarrow x \in \fff \wedge y \in \fff$
\end{frame}

@@

  \visible<2->{\begin{tikzpicture}[overlay]
    \draw[fill=white, draw=white, opacity=0.85] (-1,0) rectangle (\paperwidth,8.0);
    \node[anchor=south west,inner sep=0] (image) at (0,1.0) {\vbox{
      \only<2>{
        \centering
        \includegraphics[width=0.9\textwidth]{tierney-on-the-spectrum-of-a-ringed-topos} \\
        \footnotesize
        Miles Tierney. On the spectrum of a ringed topos. 1976.
      }

      \only<3>{
	\begin{varblock}{\textwidth}{}
	  The external meaning of
	  \[
	    \models
	      \speak{$A^\sim[X_1,\ldots,X_n]$ is anonymously Noetherian}
	  \]
	  is:
	  \medskip

@@

      \justifying
      \textbf{Proof.} The claim amounts to \mbox{$\models
      \text{``$M^\sim$}$}$\text{
      is \hil{not not} free''}$. Since~$A^\sim$ is a field, this follows from
      basic linear algebra.
    \end{column}
  \end{columns}
\end{frame}

@@

\end{frame}}

\begin{frame}[plain,c]
  \centering%
  \hil{Further reading}
  \medskip

  \framebox{\href{https://pizzaseminar.speicherleck.de/skript2/zariski-topos-klein.pdf}{\includegraphics[height=0.8\textheight]{fun-with-the-zariski-topos}}}\qquad%
  \framebox{\href{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{\includegraphics[height=0.8\textheight]{phd-cover}}}
  \par
\end{frame}

@@

    \begin{column}{0.44\textwidth}
      \begin{exampleblock}{}
        \justifying
        Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence of
        modules. If~$M'$ and~$M''$ are finitely generated, so is~$M$.
      \end{exampleblock}
    \end{column}
  \end{columns}
\end{frame}

@@

  \hil{Synthetic approach:} model schemes \hil{directly as sets} in a certain
  nonclassical set theory, the internal universe of the \mbox{\hil{big Zariski
  topos}} of a base scheme.
  \small
  \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]
\end{frame}

\fontsize{11pt}{13.2}

\begin{frame}{The big Zariski topos}
  Let~$S$ be a fixed base scheme.
  \begin{varblock}{0.9\textwidth}{Definition}
    The \hil{big Zariski topos} $\Zar(S)$ of a scheme~$S$ is equivalently
    \begin{enumerate}
      \item the topos of sheaves over~$(\Aff/S)_{\mathrm{lofp}}$,
      \item the classifying topos of local rings over~$S$ or
      \item the classifying~$\Sh(S)$-topos of local~$\O_S$-algebras which are
      local over~$\O_S$.
    \end{enumerate}
  \end{varblock}

  \begin{itemize}
    \item For an~$S$-scheme~$X$, its functor of points $\ull{X} =
    \Hom_S(\cdot,X)$ is an object of~$\Zar(S)$. It feels like \hil{the set of
    points} of~$X$.
    \item In particular, there is the ring object~$\affl$ with~$\affl(T) =
    \O_T(T)$.
    \item This ring object is a \hil{field}: nonzero implies
    invertible. \\{} [Kock 1976]
  \end{itemize}

  \jnote{1}{The objects of the category~$(\Aff/S)_{\mathrm{lofp}}$ are
  morphisms of the form~$\Spec(R) \to S$ which are locally of finite
  presentation. (Other choices of resolving set-theoretical issues of size are
  also possible.)

  A functor~$F : (\Aff/S)_{\mathrm{lofp}}^\op \to \Set$ is a sheaf
  for the Zariski topology if and only if the diagram
  \[ F(T) \to
    \prod_i F(U_i) \rightrightarrows
    \prod_{j,k} F(U_j \cap U_k) \]
  is a limit diagram for any open covering~$T = \bigcup_i U_i$ of any
  scheme~$T \in (\Aff/S)_{\mathrm{lofp}}$.

  In the case that~$S = \Spec(A)$ is affine, the big Zariski topos of~$S$ is also
  simply called ``big Zariski topos of~$A$''. It is a subtopos of the topos of
  functors $\mathrm{Alg}(A) \to \Set$ and classifies local~$A$-algebras.

  From the internal point of view of~$\Sh(S)$, the sheaf~$\O_S$ of rings is
  just an ordinary ring, and we can construct internally to~$\Sh(S)$ the big
  Zariski topos of~$\O_S$. Externally, this construction will yield a certain
  bounded topos over~$\Sh(S)$. However, as indicated on the slide, this topos
  will \emph{not} coincide with the true big Zariski topos of~$S$. To
  construct the true big Zariski topos, we have to build, internally
  to~$\Sh(S)$, the classifying topos of local \emph{and local-over-$\O_S$}
  $\O_S$-algebras.}
\end{frame}

\begin{frame}{Synthetic constructions}
  \small
  $\hil{$\boldsymbol{\mathbb{A}^n}$} = (\affl)^n = \affl \times \cdots \times \affl$

  $\begin{array}{@{}r@{}c@{}l@{}}
    \hil{$\boldsymbol{\mathbb{P}^n}$} &\phantom{}=\phantom{}& \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee
    \cdots \vee x_n \neq 0 \}/(\affl)^\times \\
    &\cong& \text{set of one-dimensional subspaces of~$(\affl)^{n+1}$} \\
    &&\qquad \text{(with~$\O(-1) = (\ell)_{\ell \? \mathbb{P}^n}$, $\O(1) =
    (\ell^\vee)_{\ell \? \mathbb{P}^n}$)}
  \end{array}$

  $\hil{$\boldsymbol{\Spec(R)}$} = \Hom_{\mathrm{Alg}(\affl)}(R, \affl) =
    \text{set of $\affl$-valued points of $R$}$

  $\hil{$\boldsymbol{TX}$} = X^\Delta$, where $\Delta = \{ \varepsilon \? \affl \,|\, \varepsilon^2 = 0 \}$

  A subset $U \subseteq X$ is \hil{qc-open} if and only if for any $x : X$
  there exist $f_1,\ldots,f_n \? \affl$ such that $x \in U \Longleftrightarrow
  \exists i\_ f_i \neq 0$.

  A \hil{synthetic affine scheme} is a set which is in bijection
  with~$\Spec(R)$ for some synthetically quasicoherent~$\affl$-algebra~$R$.

  A \hil{finitely presented synthetic scheme} is a set which can be covered by
  finitely many qc-open f.p. synthetic affine schemes~$U_i$ such that the
  intersections~$U_i \cap U_j$ can be covered by finitely many qc-open f.p.
  synthetic affine schemes.

  \jnote{1}{In the internal universe of the big Zariski topos of a base
  scheme~$S$, $S$-schemes can simply be modeled by sets (enjoying the special
  property that, in a certain precise sense, they are locally affine). This
  slide expresses some of the basic constructions of~$S$-schemes in that
  language.

>>>>

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\CC}{\mathbb{C}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LocRing}{\mathrm{LocRing}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ull}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
\newcommand{\Ll}{\vcentcolon\!\Longleftrightarrow}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}{\vdash_{\!\!\!\vec x}}
\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates

@@

  \visible<4->{
    \begin{changemargin}{-2.0em}{-0.5em}
      \begin{itemize}
        \item \ \\[-1.2em]\mbox{For any reduced ring~$A$, there is a ring~$A^\sim$ in a certain topos with}
        \[ \models \bigl(\forall x\?A^\sim\_ \neg(\exists y\?A^\sim\_ xy = 1) \Rightarrow x = 0\bigr). \]

@@

  \jnote{1}{The two displayed statements are trivial for fields. It is
  therefore natural to try to reduce the general situation to that of fields.}

@@

  Using the new reduction technique, there is a short (one-paragraph) and
  conceptual proof of Grothendieck's generic freeness lemma. It is constructive
  as a bonus; and if desired, one can unwind the resulting proof to obtain a
  constructive proof which doesn't reference topos theory. The proof obtained
  in this way is still an improvement on the previously known proofs, requiring
  no advanced prerequisites in commutative algebra, and takes
  \fixedhref{https://arxiv.org/abs/1807.01231}{about a page}.}

  \jnote{4}{Instead of passing from a given ring~$A$ to one of its
  stalks~$A_\ppp$ or quotient rings~$A/\aaa$, the reduction technique presented
  in this talk passes from~$A$ to a \emph{forcing model}~$A^\sim$.

  Unlike stalks or quotient rings, which are honest rings, the forcing
  model~$A^\sim$ is not a ring in the strict sense of the word: It doesn't
  have an underlying \emph{set} of elements, but instead an underlying
  \emph{sheaf} of elements. It is a \emph{ring object} in a certain category,
  the \emph{little Zariski topos} of~$A$. But as long as we restrict to intuitionistic
  reasoning, this difference is immaterial. Any intuitionistic theorem about
  rings applies to~$A^\sim$ as if~$A$ were a proper ring.

  Studying~$A^\sim$ is in fact the same as studying~$A$ from a certain
  different, \emph{local} point of view. The precise meaning of this statement
  will be explained on slide~3.}

  \jnote{5}{Intuitionistic logic is the same as classical logic, but without:
  \vspace*{-1em}
  \begin{itemize}
    \item the law of excluded middle: $\varphi \vee \neg\varphi$ \\[-1.5em]
    \item the law of double negation elimination: $\neg\neg\varphi \Rightarrow \varphi$ \\[-1.5em]
    \item the axiom of choice
  \end{itemize}
  \vspace*{-1em}
  If one is unfamiliar with constructive mathematics, then doing without these
  three laws seems unmotivated, rather peculiar and overtly restrictive. Here, the
  restriction to intuitionistic logic is not by some philosophical choice.
  Rather, it's by mathematical necessity. It's just a fact that in general, the
  laws of classical logic don't apply to~$A^\sim$. (Assuming a classical
  metatheory and assuming~$A$ to be Noetherian, they do iff~$A$ is of Krull
  dimension~$\leq 0$.)

  Luckily, vast amounts of commutative algebra work in the intuitionistic
  setting, as for instance evidenced by the recent 1000$^+$-page
  \fixedhref{https://arxiv.org/abs/1605.04832}{textbook by
  Lombardi and Quitté}. This claim extends to statements which are usually
  proven using maximal ideals or minimal ideal prime ideals and hence require
  Zorn's lemma. Indeed, the technique presented in this talk allows to
  constructivize some results of this kind.

  Background on constructive mathematics can for instance be found in a
  \fixedhref{https://video.ias.edu/members/1213/0318-AndrejBauer}{talk by
  Andrej Bauer}
  (\fixedhref{https://www.ams.org/journals/bull/2017-54-03/S0273-0979-2016-01556-4/S0273-0979-2016-01556-4.pdf}{written
  notes}). The standard proof that~$\sqrt{2}$ is not rational is perfectly fine
  in constructive mathematics.}
\end{frame}

@@

  \jnote{1}{In topos theory, we have lots of experience of \emph{changing
  universes} in order to \emph{force} some statements to become true. However,
  because the field condition we are aiming at is not a \emph{geometric
  sequent}, these techniques do not work here. Hence we'll take it more slowly
  and only devise a semantics which forces the given ring to be local.

  The displayed definition of a local ring is, in the presence of the axiom of
  choice, equivalent to the more common one (ring with exactly one maximal
  ideal). In constructive mathematics, the displayed definition usually works
  better.

  The key insight is that \emph{locally} (in the sense of topology/geometry),
  any ring is a local ring. That is, we may pretend that any given ring is
  local if we are prepared to pass to numerous localizations during the course
  of an argument. The semantics displayed on the next slide manages this
  localization-juggling for us.

  By~$A[f^{-1}]$, we mean the localization of~$A$ away from~$f$. This
  construction makes sense even if~$f$ is a zerodivisor, in which
  case~$A[f^{-1}]$ is the zero ring.}
\end{frame}

@@

  The four lemmas displayed on this slide, as well as all the claims on
  further slides, can be proven in very weak intuitionistic metatheories.}

  \jnote{3}{Irrespective of whether~$A$ is a local ring, its mirror
  image~$A^\sim$ is always a local ring (that is, the axioms of what it means
  to be a local ring hold under the translation rules specified by the
  semantics). A basic application of this forcing model are local-to-global
  principles. For instance:
  \begin{itemize}\justifying
    \item The statement ``the kernel of a surjective matrix over a local ring
    is finite free'' admits a constructive proof. It therefore holds
    for~$A^\sim$. Its external meaning is that the kernel of a surjective
    matrix~$M$ over~$A$ is finite locally free (there exists a partition~$1 =
    f_1 + \cdots + f_n$ such that for each~$i$, the localized
    module~$(\operatorname{ker} M)[f_i^{-1}]$ is finite free).

@@

  \jnote{1}{A topos is a special kind of category. Every topos has an
  associated \emph{internal language} which can be used to do mathematics
  \emph{internally to the topos}.

  The prototypical example of a topos is the category~$\Set$. Doing mathematics
  internally to~$\Set$ amounts to just doing mathematics in the usual sense.
  A primer on the topos-theoretic landscape is contained in
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-leipzig2018.pdf}{these
  slides}. These slides also explain the geometric reason why the intermediate value
  theorem fails in most toposes of sheaves, and why in the ef{}fective topos
  any function~$\NN \to \NN$ is computable.}

  \jnote{2-}{A free local ring~$A'$ over~$A$ is a local ring~$A'$
  together with a ring homomorphism~$A \to A'$ such that any ring
  homomorphim~$A \to R$ to a local ring uniquely factors over a local ring
  homomorphism~$A' \to R$. (A ring homomorphism is \emph{local} iff it reflects
  invertibility.)

  Assuming the Boolean Prime Ideal Theorem, one can show that there is a free
  local ring over~$A$ if and only if~$A$ has exactly one prime ideal. In this
  case~$A$ is already local, and we can take~$A' \defeq A$. If we want every
  ring to possess a free local ring over it, we need to accept ring objects of
  different toposes than~$\Set$.

  The little Zariski topos contains the \emph{generic filter} of~$A$.
  Localizing~$A$ at this filter yields the desired free local ring. It is
  precisely what was called~$A^\sim$ before. It is also known as the structure
  sheaf of~$\Spec(A)$.}

@@

    \begin{column}{0.50\textwidth}
      \small\justifying
      For a fixed ring~$R$, the localization $A' \defeq A[S^{-1}]$ with $S \defeq
      \alpha^{-1}[R^\times]$ would do the job. ($S$ is a \emph{filter}.)
      \medskip

@@

  Prime filter:\, $0 \not\in \fff$;\,\,
  $x+y \in \fff \Rightarrow x \in \fff \vee y \in \fff$;
  \hspace*{0.5pt}\,\,
  $1 \in \fff$;\,
  $xy \in \fff \Leftrightarrow x \in \fff \wedge y \in \fff$

  \jnote{1}{Any geometric theory has a \emph{classifying topos} which contains
  the \emph{generic model} of that theory (any model in any topos is uniquely
  the pullback of the generic one); if the theory under consideration is
  propositional (doesn't have any sorts), then its classifying topos can be
  chosen to be the topos of sheaves over a locale. One can also give a direct
  account of classifying locales, as a pedagogical stepping stone to the full
  theory of classifying toposes.

  The slide contains a small lie: The classical definition of the spectrum of a
  ring is via the set of prime ideals of~$A$, not filters. If the law of
  excluded middle is available, there is no difference between these
  definitions since the complement of a prime ideal is a filter and vice
  versa.
  One can also consider the classifying locale of prime ideals of~$A$.
  Its associated topological space of points is the set of prime ideals
  of~$A$ equipped with the \emph{constructible topology}.

  In an intuitionistic (but still impredicative) context, any of the
  (generalized) spaces of items~1--4 can be adopted as sensible definitions of
  the spectrum. Item~5 is then a tautology. The classical definition of
  the spectrum as a topological space doesn't work very well, because verifying
  the universal property one expects of it requires the Boolean Prime Ideal
  Theorem. Most dramatically, in some toposes there are rings which are not
  trivial yet have neither prime ideals nor filters. The classical definition
  yields in this case the empty space.}
\end{frame}

@@

  \jnote{1}{For working with~$A^\sim$, it's important to know how properties
  of~$A$ relate to properties of~$A^\sim$.

  The first displayed metatheorem justifies that, to a first approximation, the
  forcing model~$A^\sim$ is a reification of all the stalks of~$A$ into a
  single coherent entity. But crucially, this slogan is only correct for
  properties which can be put in the displayed syntactical form (called
  \emph{geometric sequents}). The reductive power of passing from~$A$
  to~$A^\sim$ results from surprising non-geometric sequents which are
  satisfied by~$A^\sim$ and not shared by~$A$, its localizations or its
  quotients.

  A slight generalization of the second metatheorem soups up a number of basic
  lemmas of algebraic geometry, there stated in geometric language. For
  instance, if~$M$ is finitely generated, then~$M^\sim$ is of finite type.
  If~$M$ is finitely presented, then~$M^\sim$ is of finite presentation. If~$M$
  is coherent, then~$M^\sim$ is coherent.}

  \jnote{2}{Surprisingly and significantly, in case that~$A$ is reduced, there
  are a number of non-geometric sequents validated by~$A^\sim$. These are
  unique features of the forcing model.

  $A^\sim$ is a field in the sense that zero is the only
  noninvertible element.

  $A^\sim$ has~$\neg\neg$-stable equality in the sense that
  \[ \models \forall x\?A^\sim\_ \forall y\?A^\sim\_ \neg\neg(x = y) \Rightarrow x = y. \]
  Classically, every set has~$\neg\neg$-stable equality; intuitionistically,
  this is a special property of some sets. It's quite useful, as some theorems
  of classical commutative algebra can only be proven intuitionistically when
  weakened by double negation. The stability then allows, in some cases, to
  obtain the original conclusion.

  $A^\sim$ is \emph{anonymously Noetherian} in the sense that any of its ideals is
  \mbox{\emph{not not}} finitely generated. A philosophically-motivated
  constructivist might be offended by this notion, since it runs counter to the
  maxim that constructive mathematics should be informative, telling us only
  that there can't not be finite generating families. However, in the forcing
  context it is a useful notion: Hilbert's basis theorem holds for it, and it
  can be put to good use in the proof of (the general case of) Grothendieck's
  generic freeness lemma.}

  \jnote{3}{The field property was already observed in the 1970s
  by Mulvey, but apparently back then its significance for applications was
  overlooked and no deeper reason for this property was known. We now know
  that it's a shadow of a forced higher-order property whose external translation
  expresses that~$A^\sim$ is quasicoherent (details are in Section~3.3 of
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
  notes}).}

  \jnote{4}{Some properties of the forcing model, which are easy to state and
  prove as properties about~$A^\sim$, have quite complex meanings when
  unravelled to refer directly to~$A$. In this way the forcing model unlocks
  observations which might otherwise be too unwieldy to manage.}

  \visible<3->{\begin{tikzpicture}[overlay]
    \draw[fill=white, draw=white, opacity=0.85] (-1,0) rectangle (\paperwidth,8.0);
    \node[anchor=south west,inner sep=0] (image) at (0,1.0) {\vbox{
      \only<3>{
        \centering
        \includegraphics[width=0.9\textwidth]{tierney-on-the-spectrum-of-a-ringed-topos} \\
        \footnotesize
        Miles Tierney. On the spectrum of a ringed topos. 1976.
      }

      \only<4>{
	\begin{varblock}{\textwidth}{}
	  The external meaning of
	  \[
	    \models
	      \speak{$A^\sim[X_1,\ldots,X_n]$ is anonymously Noetherian}
	  \]
	  is:
	  \medskip

@@

      \justifying
      \textbf{Proof.} The claim amounts to \mbox{$\models
      \text{``$M^\sim$}$}$\text{
      is \hil{not not} free''}$. Since~$A^\sim$ is a field, this follows from
      basic linear algebra.
    \end{column}
  \end{columns}

  \jnote{1}{With the baby example, we see that the new reduction technique
  presented in this talk allows to reinterpret the core content of the
  classical proof shown on slide~1 in a constructive fashion. Since the forcing
  model was set up in a constructive way, we could mechanically extract an
  explicit procedure from the new proof.

  For the second example, a remark might be in order. A basic theorem of
  undergraduate linear algebra is that any finitely generated vector space has
  a basis. In this form, the theorem cannot quite be proven intuitionistically.
  Indeed, the standard proof uses the law of excluded middle in the first step:
  \begin{indentblock}
    \scriptsize
    Let~$(x_1,\ldots,x_n)$ be a given finite generating family.
    It might be the case that one of the vectors~$x_i$ can be expressed as a
    linear combination of the others, or not.
    In the second case, the family is linearly independent and thus the vector
    space is shown to have a basis.
    In the first case, we continue with the smaller generating
    family~$(x_1,\ldots,x_{i-1},x_{i+1},\ldots,x_n)$ in an inductive fashion.
  \end{indentblock}
  Intuitionistically, the law of excluded middle is not available; however its
  double negation ($\neg\neg(\varphi \vee \neg\varphi)$) is. Threading the
  double negation through the rest of the proof yields the following theorem of
  intuitionistic linear algebra: Any finitely generated vector space is
  \emph{not not} free. It is this theorem which the proof on the slide refers
  to.}
\end{frame}

@@

  \jnote{1}{Summarizing, we can associate to any reduced ring~$A$ the forcing
  model~$A^\sim$.
  \begin{itemize}
    \item It has the pleasant property that it is a field.
    \item Reasoning about it requires that we restrict ourselves to
    intuitionistic logic.
  \end{itemize}
  Depending on the application, this trade-off can not be useful at all, or be
  quite valuable. It certainly is so for proving Grothendieck's generic
  freeness lemma, simplifying a multi-page argument to a single and conceptual
  paragraph. (The previous slide only showed a fragment of the general
  statement of Grothendieck's generic freeness lemma. The technique outlined in
  these slides is amenable to the full version. Details are in Section~11.5 of
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
  notes}.)

  The ideas underlying this new reduction technique can also be used for
  different purposes: for understanding, in a rigorous way, notions of algebraic
  geometry as notions of algebra, and for developing a synthetic account of
  algebraic geometry. This is very briefly touched upon on the next two
  slides.}

  \jnote{2}{
    \centering
    \hil{Further reading}
    \medskip

    \framebox{\fixedhref{https://pizzaseminar.speicherleck.de/skript2/zariski-topos-klein.pdf}{\includegraphics[height=0.8\textheight]{fun-with-the-zariski-topos}}}\qquad%
    \framebox{\fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{\includegraphics[height=0.8\textheight]{phd-cover}}}
    \par
  }
\end{frame}}

@@

    \begin{column}{0.44\textwidth}
      \begin{exampleblock}{}
        \justifying
        Let~$0 \to M' \to M \to M'' \to 0$ be a short exact sequence of
        modules. If~$M'$ and~$M''$ are finitely generated, so is~$M$.
      \end{exampleblock}
    \end{column}
  \end{columns}

  \jnote{1}{One doesn't need to be an expert in topos theory in order to know
  that many notions in algebraic geometry are inspired by notions in algebra
  and that proofs in algebraic geometry often proceed by reducing to algebra.

  If~$X$ is a scheme, the internal language of the topos~$\Sh(X)$ is a way of
  making this connection precise: In many cases, the former are simply
  interpretations of the latter internal to~$\Sh(X)$. Because this connection
  is precise instead of informal, additional value is gained: We can skip many
  basic proofs in algebraic geometry because they're just externalizations of
  proofs in algebra carried out internally to~$\Sh(X)$.}

  \jnote{2}{A basic example is as follows. A short exact sequence of sheaves of
  modules looks like a short exact sequence of plain modules from the internal
  point of view of~$\Sh(X)$. If the two outer sheaves are of finite type, then
  from the internal point of view, the two outer modules will look like
  finitely generated modules. Because the standard proof of the proposition
  quoted on the lower right is intuitionistically valid, it follows that, from
  the internal point of view, the middle module is too finitely generated.
  Consulting the dictionary a second time, this amounts to saying that the
  middle sheaf is of finite type.

  A more advanced example is: The theorem ``any finitely generated vector space
  does \emph{not not} have a basis'' of constructive linear algebra entails, by
  interpretation in~$\Sh(X)$, that any sheaf of finite type over a reduced
  scheme is finite locally free on a \emph{dense open subset}.

  More details on this research program can be found in
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
  notes}, partly reported on at the
  \fixedhref{https://www.youtube.com/watch?v=7S8--bIKaWQ}{2015 IHÉS topos
  theory conference}. Even though many important dictionary entries are still
  missing (for instance pertaining to derived categories and intersection
  theory), I believe that it is already in its current form useful to working
  algebraic geometers.

  The next slide illustrates a further, different way of approaching algebraic
  geometry using topos theory.}
\end{frame}

@@

  \jnote{1}{At the
  \fixedhref{https://sbseminar.wordpress.com/2009/08/06/algebraic-geometry-without-prime-ideals/}{Secret
  Blogging Seminar}, there was an insightful long-running discussion on the
  merits of the two approaches to algebraic geometry. Two disadvantages of the approach using
  locally ringed spaces is that the underlying topological spaces don't
  actually parametrize ``honest'', ``geometric'' points, but the more complex
  notion of irreducible closed subsets; and that they don't work well in a
  constructive setting. (For this, they would have to be replaced by locally
  ringed locales.)

  The functorial approach is more economical, philosophically rewarding, and
  works constructively. Given a functor~$F : \Ring \to \Set$, we imagine~$F(A)$
  to be the set of~``$A$-valued points'' of the hypothetical scheme described
  by~$F$, the set of ``points with coordinates in~$A$''. These sets have direct
  geometric meaning. However, typically only field-valued points are easy
  to describe. For instance, the functor representing projective~$n$-space is
  given on fields by
  \[ \begin{array}{r@{}c@{}l}
  K &{}\longmapsto{}& \text{the set of lines through the origin in~$K^{n+1}$} \\
  && \qquad \cong \{ [x_0 \hg \cdots \hg x_n] \,|\, \text{$x_i \neq 0$ for some~$i$} \},
  \end{array} \]
  whereas on general rings it is given by
  \[ \begin{array}{r@{}c@{}l}
    A &{}\longmapsto{}& \text{the set of quotients~$A^{n+1} \twoheadrightarrow P$,
    where~$P$ is projective,} \\
    && \qquad \text{modulo isomorphism}.
  \end{array} \]
  It is these more general kinds of points which impart a sense of
  cohesion on the field-valued points, so they can't simply be dropped from
  consideration.\bigskip}

  \hil{Synthetic approach:} model schemes \hil{directly as sets} in a certain
  nonclassical set theory, the internal universe of the \mbox{\hil{big Zariski
  topos}} of a base scheme.
  \small
  \[ \{ (x,y,z) \? (\affl)^3 \,|\, x^n+y^n-z^n=0 \} \]

  \jnote{2}{This tension is resolved by observing that the category of
  functors~$\Ring \to \Set$ is a topos (the \emph{big Zariski topos} of~$\Spec(\ZZ)$)
  and that we can therefore employ its internal language. This language takes
  care of juggling stages behind the scenes. For instance, projective~$n$-space
  can be described by the naive expression
  \[ \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee \cdots \vee x_n
  \neq 0 \}/(\affl)^\times. \]
  This example illustrates the goal: to develop a synthetic account of
  algebraic geometry, in which schemes are plain sets and morphisms between
  schemes are maps between those sets. It turns out that there are many
  similarities with the well-developed synthetic account of differential
  geometry, but also important differences, and it also turns out that
  synthetic algebraic geometry has close connections to a certain age-old
  unsolved problem in topos theory, the \emph{mystery of nongeometric
  sequents}.

  Details are in
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/slides-como2018.pdf}{this
  set of slides}.}
\end{frame}

\backupend
